Typed Functional Genetic Programming

### 2013-07-28 tex/thesis.tex
tex

\newcommand{\sexprTree}{sexpr-tree\xspace}
\newcommand{\atTree}{@-tree\xspace}

@@

\subsection{Benefits of generating \lterms in \lnf}

@@

It is parameterized by parametr $q \in (0,1)$.

@@

\chapter{Problems}
	In this section will be presented usage of the system in order to solve specific problems.

		\section{Even Parity Problem}
		\section{Big Context}
		\section{Fly}
		\section{Simple Symbolic Regression}
		\section{Artificial Ant}
		\section{Boolean Alternate}

>>>>

\newcommand{\sexprTree}{sexpr-tree\xspace}
\newcommand{\SexprTree}{Sexpr-tree\xspace}
\newcommand{\atTree}{@-tree\xspace}

@@

\subsection{Benefits of generating \lterms in \lnf}
\label{benefits}

@@

It is parameterized by parametr $q \in (0,1)$, its default well-performing 
value is $q = 0.75$. \\

@@

\chapter{Problems}
	In this section will be presented usage of the system in order to solve specific problems. Several problems will be described and some of them compared with results
of others.

\section{Even-parity problem}
The even-parity function is a boolean function taking as inputs $N$
boolean values and returning \textit{True} if an even number of inputs 
are \textit{True}. For odd number it returns \textit{False}.

The reason why examine this problem is that it has been used by many researchers
as a benchmark for GP \cite{todo,yu01}.

We compare our results with that obtained by Yu in \cite{yu01},
where is presented approach to evolve recursive and modular programs
by use of higher-order functions and $\lambda$-abstractions.
Since the results are highly sensitive to set of \textit{building blocks}, 
we use very similar set of \textit{building blocks} in order to make 
the results comparable.

\begin{align*}
\sigma = [Boo&l] \ar Bool\\
\Gamma = \{
  and   &: Bool \ar Bool \ar Bool                              ,\\
  or    &: Bool \ar Bool \ar Bool                              ,\\
  nand  &: Bool \ar Bool \ar Bool                              ,\\
  nor   &: Bool \ar Bool \ar Bool                              ,\\
  foldr &: (Bool \ar Bool \ar Bool) \ar Bool \ar [Bool] \ar Bool ,\\
  head' &: [Bool] \ar Bool                                   ,\\
  tail' &: [Bool] \ar [Bool]                              \}\\
\end{align*}

The type $[Bool]$ stands for \textit{list of $Bool$s} and for purpose of
this problem is considered atomic.

Unlike in \cite{yu01}, we use specific instance of polymorphic 
function \texttt{foldr}. 
And modifications of functions \texttt{head} 
(returning the first element of the list) 
and \texttt{tail} (returning the list without the first element) are used; 
making them total by returning default value \textit{False}
and \texttt{[]}, respectively.

The \texttt{fold} function is higher order function widely used in \textit{Haskell}
programming language, recursively defined as follows.

\texttt{~\\
foldr :: (a -> b -> b) -> b -> [a] -> b \\
foldr f z []     = z \\
foldr f z (x:xs) = f x (foldr f z xs)  
}\\

\Lets demonstrate \texttt{foldr}'s behavior on the following simple example.

\texttt{~\\
foldr xor False [True,False,True] \\
= xor True (foldr False [False,True]) \\
= xor True ( xor False ( foldr False [True] ) )\\
= xor True ( xor False ( xor True (foldr False [] ) )\\
= xor True ( xor False ( xor True False) )\\
= False
}\\

The reasons why is this difficult problem 
for GP to solve are following \cite{todo}:

\begin{itemize}
 \item Since a single change of one input generates a different output,
       the output is highly sensitive to all inputs.
 \item The building blocks set does not contain functions \textit{xor}
       or \textit{eq}, which are keys to solving the problem.
       But those two functions may be discover by GP.  
\end{itemize}

We use the same fitness function as in \cite{yu01}. 
The fitness function examines the individual by giving
it all possible boolean lists of length 2 and 3.
Thus resulting in $2^2 + 2^3 = 12$ test cases.
For each correct response it receives $1/12$, 
therefore correct solution receives fitness value $1$.\\

\begin{todo}
 & zmínit tu, že to při týhle building block stačí na general solution...
\end{todo}

\subsection{Experiments}

Series of five experiments have been performed to test properties of our system.
Each experiment consisted of 50 independent runs of GP algorithm.
Each run had maximally 50 generations
(50 + 1 for generation 0) and 500 individuals as population size.\\

In those experiments we examined influences of the following factors:

\begin{itemize}
 \item Option of preserving the best individual into the next population.
 \item Option of performing $\eta$-normalization of generated individuals 
      (conversion from \lnf to \benf, see \ref{benefits}).
 \item Representation of individuals: \sexprTree vs \atTree.
 \item Option of performing optimized abstraction elimination
       (see \ref{toSKI}) instead of the simple one.
\end{itemize}

We analyze the ability of the system to produce a correct solution.
We are interested in percentage of runs that led to discovery of
correct solution. And we also use the popular measurement  
method within GP field --- the \textit{performance curves}
described in \cite{koza92}.

First performance curve is the cumulative \textit{probability of success}
$P(M,i)$, where $i$ is generation number ($i = 0$ for initial population) 
and $M$ is the size of the population. 
$P(M,i)$ is probability that a correct solution will
be find in one run of GP with maximal number of generations $i$ and 
population size $M$. $P(M,i)$ is computed as percentage of 
runs yielding a correct solution in generation $i$ or earlier.

From values of $P(M,i)$ are computed values of the second performance curve
$I(M,i,z)$
--- the total number of individuals that must be processed to yield a correct
solution with probability $z$ by multiple runs of GP with maximal 
number of generations $i$ and population size $M$ (usually $z = 0.99$).

As is described in \cite{koza92},
The number $R(M,i,z)$ of independent runs required to yield a correct solution 
is computed as

$$ R(M,i,z) = \lceil\frac{\log(1-z)}{\log(1-P(M,i))}\rceil ,$$

thus $I(M,i,z)$ is computed as

$$I(M,i,z) = (i+1) \cdot M \cdot R(M,i,z) .$$\\

The setup of the first experiment was following:\\

\begin{enum}
 & \textit{Geometric} search strategy for generating individuals with default $q=0.75$.
   This option is unchanged for all five experiments.
 & No best individual preservation.
 & $\eta$-normalization was not performed.
 & Individuals represented as \sexprTree{}s. 
 & Simple abstraction elimination.
\end{enum}~

Thus the simplest one setup; all extra features are off. 
In each following experiment one extra feature will be turned on,
in hope that it will enhance the performance.

The correct individual was yield 
in 17 out of 50 runs, i.e., 34\% success rate.
The performance curve for $P(M,i)$ is following.

\includegraphics[scale=0.65]{reports/ep/report1/probabs.png}

Interestingly, in 2 runs out of 50, a correct solution was present in the initial population.

The performance curve for $I(M,i,z)$ is following.

\includegraphics[scale=0.65]{reports/ep/report1/indivs.png}

The smallest value in this graph is used to indicate the minimum effort
(i.e. number of individuals to be processed) for GP to yield a correct solution.

Thus for this experiment this analysis suggest
to perform only term generating and no evolution at all, 
because the smallest value 56500 is in generation 0.

This experiment took 59 minutes (on average desktop computer).\\

Second experiment was different from the first one in that it 
preserved the best individual to the next generation.

It scored 22/50, i.e., 40\% success rate. A correct solution was
created once in the initial generation and again it was
sufficient to make the value in generation 0
be the smallest value in the $I(M,i,z)$ curve; now 114000
 individuals to be processed.
This experiment also took 59 minutes.\\

Third experiment was different from the second one in that 
$\eta$-normalization of generated individuals was performed.

It scored 27/50, i.e., 54\% success rate. 
Two runs contained a correct solution in the initial population.
But now the $I(M,i,z)$ was best for generation 1 with 56000
 individuals to be processed.
This experiment took 53 minutes.\\

Fourth experiment was different from the third one in that
\atTree representation was used instead of \sexprTree.

It scored 34/50, i.e., 68\% success rate. Again two runs
contained a correct solution in the initial population.
The $I(M,i,z)$ was best for generation 0 with 56500 
individuals to be processed.
This experiment took 64 minutes.

As we will see, this setup was the most successful one,
thus we again show the performance curves.

\includegraphics[scale=0.65]{reports/ep/report4/probabs.png}

\includegraphics[scale=0.65]{reports/ep/report4/indivs.png}

Fifth, last experiment was different from the fourth one in that
optimized abstraction elimination was used instead of the simple one.

It scored 33/50, i.e., 66\% success rate. One run
contained a correct solution in the initial population.
The $I(M,i,z)$ was best for generation 6 with 59500 
individuals to be processed.
This experiment took only 30 minutes. \\

\subsection{Comparison with other results}

Results presented by Yu \cite{yu01} are better then our results;
they scored 40/50, i.e., 80\% success rate.  
Her $I(M,i,z)$ was best for generation 4 with 17500 
individuals to be processed.
Yu also uses crossover as the only genetic operator, thus we
can narrow the discussion to discussion of generating method and 
crossover method.

When we observe results of our experiments, it seems
that our system is pretty good in term generation, but not so good 
in crossover. Since the main focus point of this thesis is
the term generation, it is positive rather then negative observation.

Design of our crossover comes from that we wanted it to be
very simple generalization of standard tree-swapping crossover.
Whereas the crossover used by Yu is more structure oriented
--- as is stated in \cite{yu01} the crossover \textit{can only operate 
between nodes in two main programs or between nodes in the same
kind of $\lambda$ abstraction (i.e., $\lambda$ abstractions that 
represent the same function arguments to the same higher order function).}
%This seems to be great advantage for optimizing the

Since our method has better success rate for generation 0, it would 
be interesting to try combination of our generating 
method with crossover used by Yu.\\

Positive observation is that our result at least outperforms 
other results mentioned in \cite{yu01} with which her method is compared.

\textit{Generic genetic programming} scored 17/60, i.e. 28\%; 
minimum $I(M,i,z) = 220,000$.

\textit{GP with ADFs} scored 10/29, i.e. 34\%; 
minimum $I(M,i,z) = 1,440,000$.

\begin{todo}
& napsat tam jak vypadaj nějaký nalezeneý řešení
\end{todo}

\section{Flies and Apples}

\textit{Flies and Apples} is problem consisting in breeding 
control programs for a fly agent in a simple world.

Fly world is a square grid. 
Each cell is ether empty or contains precisely one object.
There are three kinds of objects:

~\begin{enum}
 & Wall
 & Apple
 & Fly
\end{enum}~

Apples and flies have \textit{energy}.

Moreover, each fly has its control program and inner state, 
so called \textit{registers}.

There is a queue of all flies in the fly world determining which 
fly is next to make an action.
After the fly in the front of the queue performs its action, 
it is put to the back. In other words, flies are cyclically
taking turns similarly like in common card games.\\

Fly can perform two kinds of action:

~\begin{enum}
 & \textit{Trave.} Moves the fly to one of four adjacent cells determined by 
   direction specified in the action.  
   Result of the travel action depends on the content of the target cell.    
   && If it is empty, then the fly simply moves to that cell.
   && If it contains wall, then no movement is performed.
   && If it contains an apple, then the fly moves to that cell, the apple 
      is eaten and its energy is added to energy of the fly.  
   && If it contains another fly, then the fly moves to that cell, the fly with bigger 
      energy stays alive and the weaker one is eaten
      and its energy is added to energy of the stronger fly.  
 & \textit{Split.} Puts a daughter fly to one of four adjacent cells.
    Together with the direction, amount of energy and new registers 
    for the fly are specified by this action. The mother gives this energy to 
    its child. In order to be successful, the target cell must be empty
    and the mother fly must have energy greater then 1. 
\end{enum}~

\section{Simple Symbolic Regression}

\section{Artificial Ant}

\section{Big Context}

\section{Boolean Alternate}
